---
title: "Assignment 3"
author: "Hussain Abdul Kadhir"
date: last-modified
format:
  pdf:
    fontsize: 8pt
---

## Data Description:
The data set “p4.20” in MPV package, discusses an experiment to determine the influence of the following five factors on an appropriate measure of the whiteness of rayon ($y$).

-   $x_1$: acid bath temperature
-   $x_2$: cascade acid concentration
-   $x_3$: water temperature
-   $x_4$: sulfide concentration
-   $x_5$: amount of chlorine bleach

::: {.callout-note}
## Questions:
a. Create the scatterplot of each independent variable versus the dependent
variable. Which variables shows a possible linear relationship?
b. Discuss and find the best regression model.
c. Perform a residual analysis including the residual plots. Discuss about the validity of regression model assumption, like normality and constant variance.
d. Based on the residual analysis in Part c, find possible outlier
:::

```{r setup, warning=FALSE, message=FALSE}
library(MPV)
```

```{r data}
data(p4.20)
head(p4.20)
summary(p4.20)
attach(p4.20)
```

## a. Scatterplots and correlation
```{r scatter_plots_and_correlation, fig.width = 4, fig.height = 3, fig.align = "center"}

# scatter plots 
plot(acidtemp, y, main="acidtemp vs y")
plot(acidconc, y, main="acidconc vs y")
plot(watertemp, y, main="watertemp vs y")
plot(sulfconc, y, main="sulfconc vs y")
plot(amtbl, y, main="amtbl vs y")

# correlation 
cor(y, acidtemp)
cor(y, acidconc)
cor(y, watertemp)
cor(y, sulfconc)
cor(y, amtbl)
# acidtemp shows potential linear relationship with response y
```

## b. Model selection
```{r simple_regression, tidy=TRUE}



# better way to do this combination using loop
# we need one loop for different number of variables and another loop for different combinations of variables
vars <- c("acidtemp", "acidconc", "watertemp", "sulfconc", "amtbl")
fits <- list()
r_squared <- c()
adj_r_squared <- c()
counter <- 1 # since two loops, to keep track of the fit number
for(i in 1:length(vars)){
  comb <- combn(vars, i)
  for(j in 1:ncol(comb)){
    formula <- as.formula(paste("y ~", paste(comb[, j], collapse = " + ")))
    fits[[counter]] <- lm(formula)
    r_squared[counter] <- summary(fits[[counter]])$r.squared
    adj_r_squared[counter] <- summary(fits[[counter]])$adj.r.squared
    counter <- counter + 1
  }
}

# index of max r-squared
(which.max(r_squared))
# index of max adj r-squared
(which.max(adj_r_squared))


# data frame to store the results
model_results <- data.frame(
  model = sapply(fits, function(x) paste(names(coef(x))[-1], collapse = " + ")),
  fit_i = 1:length(fits),
  r_squared = r_squared,
  adj_r_squared = adj_r_squared
)

# sorting the data frame by adj r-squared using dplyr
library(dplyr)
head(model_results <- model_results %>%
  dplyr::arrange(desc(adj_r_squared)))
fits[[6]]$model[0,] # model variables

summary(fits[[16]])
summary(fits[[27]])
summary(fits[[6]])
summary(aov(fits[[6]]))
# fit 6 is the optimum based on r-squared and adj r-squared values, as # the overall model is significant, 
# although acidtemp is significant at 0.05 level, but acidconc is not significant at 0.05 level, but significant at 0.1 level


# AIC is a measure of the relative quality of a model 
# It is based on the likelihood function
# penalizes models with more parameters to prevent overfitting. 
# The model with the lowest AIC value is better
# we can check AIC of the models
aic_values <- sapply(fits, AIC)
# sapply is used to apply the AIC function to each model in the fits list and return a vector of AIC values
which.min(aic_values)
# fit 6 has the lowest AIC, so we can consider it as the best model

# using built-in stepwise regression 
# which uses AIC as the criteria for model selection
step(lm(y ~ acidtemp + acidconc + watertemp + sulfconc + amtbl), direction = "both")

# we end up with the same model, which is y ~ acidtemp + acidconc

fit <- fits[[6]]
```

### Check the model with out built in function
```{r multiple_regression_without_built_in_function, tidy=TRUE}

# y ~ acidtemp + acidconc with out built in function

(n  <- length(y)) # data points
(p <- 2) # variables


# setting up the X matrix
X = matrix(rbind(rep(1, n), acidtemp, acidconc), byrow = T, ncol = p+1) 
dim(X)

# beta_hat ~ parameters
(beta_hat = solve(t(X)%*%X)%*%t(X)%*%y)
fit$coefficients # from lm function


# y_hat ~ response
y_hat = X%*%beta_hat
head(y_hat) # from our computation
head(fit$fitted.values) # from lm function

# e_hat ~ error
e = y - y_hat
head(e) # from our computation
head(fit$residuals) # from lm function


#ms_residual 
ss_residual <- (t(y)%*%y - t(beta_hat)%*%t(X)%*%y)
(ms_residual <- ss_residual/(length(y) - (ncol(X))))
summary.aov(fit)[[1]]["Residuals", "Mean Sq"] # from lm function

#ms_regression
ss_regression <- t(beta_hat)%*%t(X)%*%y-(sum(y)^2/n)
(ms_regression <- ss_regression/(ncol(X)-1))
(summary.aov(fit)[[1]]["acidtemp", "Mean Sq"] + summary.aov(fit)[[1]]["acidconc", "Mean Sq"])/2 # from lm function

#ms_total
ss_total <- (t(y)%*%y-(sum(y))^2/n)
ms_total <- ss_total/(n-1)


# testing the overall significance of the model
# H_0 : beta1 = beta2 = 0
# H_1 : at least one beta is not equal to zero
(f_0 <- ms_regression/ms_residual)
(summary.aov(fit)[[1]]["acidtemp", "F value"] + summary.aov(fit)[[1]]["acidconc", "F value"])/2 # from lm function

p_value <- 1 - pf(f_0, df1 = ncol(X)-1, df2 = length(y) - ncol(X))
p_value < 0.05 # significant at 0.05 level



# testing individual significance of the coefficients


# for acidtemp
# H_0 : beta1 = 0
# H_1 : beta1 != 0
cjj <- diag(solve(t(X)%*%X))
se_beta1 <- sqrt(ms_residual*cjj)
(t_0_acidtemp <- beta_hat[2]/se_beta1[2])
summary(fit)$coefficients["acidtemp", "t value"] # from lm function


# for acidconc
# H_0 : beta2 = 0
# H_1 : beta2 != 0
(t_0_acidconc <- beta_hat[3]/se_beta1[3])
summary(fit)$coefficients["acidconc", "t value"] # from lm function


# r-squared
(r_sq <- ss_regression/ss_total)
summary(fit)$r.squared # from lm function


# adj r-squared
(adj_r_sq <- 1 - ms_residual/ms_total)
summary(fit)$adj.r.squared # from lm function
```


::: {.callout-note}

## Interpretation of the model:
-   Both `acidtemp` and `acidconc` are significant at 0.05 level, since their
    t-statistics are greater than the critical value of t with n-p degrees of freedom.
-   The overall model is significant, since the p-value for the F statistic is less than 0.05.
-   The R-squared value of the model is 0.422, which means that 42.2% of the variability in the response variable can be explained by the model.
-   The adjusted R-squared value is 0.372, which is a more accurate measure of the goodness of fit of the model, since it takes into account the number of predictors in the model.
-   Since both the overall model and the individual coefficients are significant, we can conclude that there is a significant linear relationship between the response variable and the predictors (acidtemp and acidconc).
-   But the model doesn't explain variability in the response variable very well, since the R-squared value is not very high.
:::


## c. Residual analysis
```{r residual_analysis, fig.width = 4, fig.height = 3, fig.align = "center"}
# residuals are deviation between data and fit
e = y - y_hat

# mean of residuals should be around zero
round(mean(e), 5)

# variance of residuals can be estimated by ms_residual
# rss <- t(e) %*% e
var(e) # here rss / (n - 1) 
round(ms_residual, 5) # here rss / (n - p) since p = 3
# ms_residual is an unbiased estimator of the variance of the error term, sigma^2

## scaling residuals

# 1 standardized residuals

d_res <- e/sqrt( as.numeric(ms_residual))

head(d_res)
# basically we now have variances of the residuals to be 1
# outliers can be checked with
sum(abs(d_res) > 3)

# we can plot 
hist(d_res, breaks = 15)

# no outliers since we have no standardized residuals greater than 3 in absolute value

#  based on how we estimated the variance of the residuals, we ca have other types of residuals

# 2 studentized residuals
r_res <- e/sqrt( as.numeric(ms_residual) * (1 - hatvalues(fit)) )
# or rstandard(fit2)
sum(abs(r_res) > 3)

# 3 R - student residuals
# removing the i-th observation and re-estimating the variance of the residuals
s_sq <- numeric(length(y))
for (i in 1:length(y)) {
  X_i <- X[-i, ]
  y_i <- y[-i]
  beta_i <- solve(t(X_i) %*% X_i) %*% t(X_i) %*% y_i
  e_i <- y_i - X_i %*% beta_i
  s_sq[i] <- t(e_i) %*% e_i / (length(y_i) - ncol(X_i))
}

t_res <- e/sqrt( s_sq * (1 - hatvalues(fit)) )
# or rstudent(fit2)
sum(abs(t_res) > 3)

# 4 press residuals
p_res <- e/(1 - hatvalues(fit))



# Diagnostic plots
# residuals vs fitted values for checking constant variance and finding outliers
plot(fitted.values(fit), resid(fit), main = "Residuals vs Fitted")
plot(fitted.values(fit), d_res, main = "Standardized Residuals vs Fitted")
points(fitted.values(fit), d_res, col = ifelse(abs(d_res) > 3, "red", "black"), pch = 1)
plot(fitted.values(fit), r_res, main = "Studentized Residuals vs Fitted")
points(fitted.values(fit), r_res, col = ifelse(abs(r_res) > 3, "red", "black"), pch = 1)
plot(fitted.values(fit), t_res, main = "R-Student Residuals vs Fitted")
points(fitted.values(fit), t_res, col = ifelse(abs(t_res) > 3, "red", "black"), pch = 1)
plot(fitted.values(fit), p_res, main = "Press Residuals vs Fitted")
points(fitted.values(fit), p_res, col = ifelse(abs(p_res) > 3, "red", "black"), pch = 1)

# residuals vs independent variables for checking constant variance
plot(acidconc, resid(fit), main = "Residuals vs acidconc")
plot(acidtemp, resid(fit), main = "Residuals vs acidtemp")
# qq plot for normality
qqnorm(resid(fit), main = "QQ Plot of Residuals")
qqline(resid(fit))

# test for normality
shapiro.test(resid(fit))
# since p-value is greater than 0.05, we fail to reject the null hypothesis
# which means the residuals are normally distributed
```

::: {.callout-note}

## Model assumptions and outliers:
-   There is a funnel out behaviour in the residuals vs fitted values plot, which indicates that the variance of the residuals is not constant. This violates the homoscedasticity assumption of linear regression. This is same for all scaled residuals.
-   The qq plot of the residuals shows that the residuals are approximately normally distributed, which satisfies the normality assumption of linear regression.
-   Shapiro-Wilk normality test also supports the normality of the residuals, since the p-value is greater than 0.05.
-   The residuals vs independent variables plots doesnt clearly show any pattern, but there is some indication of non-constant variance.
-   The `d_res`, `r_res`, and `t_res` plots do not show any outliers, since there are no points with absolute values greater than 3.
-   The `p_res` plot shows some points with large press residuals, which may indicate influential observations. However, since there are no standardized or studentized residuals greater than 3 in absolute value, we do not have any outliers in the model.
:::
