---
title: "Worksheet - 1"
author: "Hussain Abdul Kadhir"
date: last-modified
format:
  pdf:
    fontsize: 8pt
---
## Dataset II: Table B.17 Patient Satisfaction Data
```{r setup, warning=FALSE, message=FALSE, fig.width = 4, fig.height = 3, fig.align = "center"}
library(MPV)
data(table.b17)
head(table.b17)
attach(table.b17)

# Satisfaction is the dependent variable
(n <- length(Satisfaction))

# renaming the variables for easy coding
y <- Satisfaction
x1 <- Age
x2 <- Severity
x3 <- Surgical.Medical
x4 <- Anxiety

# scatter plots
plot(y ~ x1 + x2 + x3 + x4)
cor(x1, y) # highest
cor(x2, y)
cor(x3, y)
cor(x4, y)
```

:::::: {.callout-note}

# Scatter plot and correlation
-   From the scatter plots `x1` is showing negative relationship with dependent variable (correlation = -0.8707049)
-   Rest of the variables are not showing visible linear relationship, which is further seen in the correlation values
:::
```{r finding_the_best_model}

fit1 <- lm(y ~ x1)
fit2 <- lm(y ~ x2)
fit3 <- lm(y ~ x3)
fit4 <- lm(y ~ x4)
fit5 <-lm(y ~ x1 + x2)
fit6 <-lm(y ~ x1 + x3)
fit7 <-lm(y ~ x1 + x4)
fit8 <-lm(y ~ x2 + x3)
fit9 <-lm(y ~ x2 + x4)
fit10 <-lm(y ~ x3 + x4)
fit11 <-lm(y ~ x1 + x2 + x3)
fit12 <-lm(y ~ x1 + x2 + x4)
fit13 <-lm(y ~ x1 + x3 + x4)
fit14 <-lm(y ~ x2 + x3 + x4)
fit15 <-lm(y ~ x1 + x2 + x3 + x4)

r_squared <- c()
adj_r_squared <- c()
for(i in 1:15){
  r_squared[i] <- (summary(get(paste0("fit", i)))$r.squared)
}
for(i in 1:15){
  adj_r_squared[i] <- summary(get(paste0("fit", i)))$adj.r.squared
}
r_squared
adj_r_squared
max(r_squared)
max(adj_r_squared)

# fit5 has largest adj_rsquared
fit <- fit5
summary(fit)
summary(aov(fit))
```


:::::: {.callout-note}
# Model selection
-   `fit5` i.e. `y ~ x1 + x2` is the best based adjusted r-squared (`Multiple R-squared:  0.8096,	Adjusted R-squared:  0.7923 `)
-   Overall model is significant (`F-statistic: 46.77 on 2 and 22 DF,  p-value: 1.193e-08`)
-   Both the variables `x1` and `x2` are significant as well.
:::

```{r model_diagnostics, fig.width = 4, fig.height = 3, fig.align = "center"}

plot(fit$fitted.values, fit$residuals, main = "Residual vs fitted")
plot(fit$fitted.values, rstandard(fit), main = "Residual vs fitted")
abline(3,0, lty = 2, col = "red")
abline(-3,0, lty = 2, col = "red")
plot(x1, fit$residuals, main = "Residual vs x1")
plot(x2, fit$residuals, main = "Residual vs x2")
plot(x3, fit$residuals, main = "Residual vs x3")
plot(x4, fit$residuals, main = "Residual vs x4")
plot(fit)
(abs(rstandard(fit)) > 3)
(abs(rstandard(fit)) > 3)[17]
sum(abs(rstandard(fit)) > 3)
sum(abs(rstudent(fit)) > 3)
qqnorm(fit$residuals)
qqline(fit$residuals, lty = 2, col = "red")
shapiro.test(fit$residuals)
```

:::::: {.callout-note}
## Diagnostics
-   Overall model is not showing any funnel shape
-   $17^{th}$ is the observation with large residual based on both `rstandard` and `rstudent` residuals
-   QQplot also shows normality, with $17^{th}$ showing deviation
-   Normality of residuals is not rejected by Shapiro-Wilk test (`p-value = 0.06411`)
:::


```{r removing_17th_observation}
data(table.b17)
df <- table.b17
df <- df[-17,]
attach(df)
y <- Satisfaction
x1 <- Age
x2 <- Severity
fit_updated <- lm(y ~ x1 + x2)
summary(fit_updated)
summary(fit)
summary(aov(fit))

summary(fit)$adj.r.squared
summary(fit_updated)$adj.r.squared
```

:::::: {.callout-note}
## Results after updating the model
-   After removing the outlier($17^{th}$ observation), the adjusted r-squared jumped from `0.79` to `0.89`.
-   This clearly shows the influence of the outlier
-   Overall conclusion: the varibles `age` and `severity` can explain `89%` of the variability. The model's assumptions are all met
:::