---
title: "Worksheet - 2"
author: "Hussain Abdul Kadhir"
date: last-modified
format:
  pdf:
    fontsize: 8pt
    fig.width: 4
    fig.height: 3
    fig.align: "center"
---


```{r }
#| warning: false
library(MPV)
library(knitr)
data(p7.6)

kable(p7.6)

y <- p7.6$y
x1 <- p7.6$x1
x2 <- p7.6$x2

p1 <- plot(x1, y)
p2 <- plot(x2, y)
# correlation
cor(x1, y)
cor(x2, y)
```

## Initial analysis

-    `x2` is showing some positive relationship with `y` (correlation = 0.98) while `x1` is just showing weakrelationship (correlation = 0.56)


```{r fits}
#a 
model1 <- lm(y ~ x1 + x2)
model2 <- lm(y ~ x1 + x2 + I(x1^2) + I(x2^2))

summary(model1)
summary(model2)
```


# Model 1

```{r normality_plot_model_!}
# normality plot
qqnorm(residuals(model1))
qqline(residuals(model1))
```

```{r normality}
shapiro.test(residuals(model1))
```

## Normality of Residuals

-   Virtually all the points are on the line, which indicates that the residuals are normally distributed.
-   Additionally, shapiro test also confirms the normality of residuals (p-value = 0.871)




```{r residual_plot}
# residual vs fitted plot
preds <- predict(model1)
plot(preds, residuals(model1))
abline(h = 0, col = "red")
```


```{r student_residuals}
# rstudent residuals
rstudent_residuals <- rstudent(model1)
plot(preds, rstudent_residuals)
abline(h = 0, col = "red")
```

```{r studentized_residuals}
# studentized residuals
studentized_residuals <- rstandard(model1)
plot(preds, studentized_residuals)
abline(h = 0, col = "red")
```

## Interpretation of residual plots for model1
-    All three plots are not showing any funnel shape, which indicates that the variance of residuals is constant across fitted values.
-    There are no points with large residuals (more than 3 standard deviations), which indicates that there are no outliers 


# Model 2

```{r }
# normality plot
qqnorm(residuals(model2))
qqline(residuals(model2))
```

```{r }
shapiro.test(residuals(model2))
```

## Normality of Residuals

-   Virtually all the points are on the line, which indicates that the residuals are normally distributed.
-   Additionally, shapiro test also confirms the normality of residuals (p-value = 0.87)




```{r }
# residual vs fitted plot
preds <- predict(model2)
plot(preds, residuals(model2))
abline(h = 0, col = "red")
```


```{r student_residuals_M2}
# rstudent residuals
rstudent_residuals <- rstudent(model2)
plot(preds, rstudent_residuals)
abline(h = 0, col = "red")
```

```{r studentized_residuals_M2}
# studentized residuals
studentized_residuals <- rstandard(model2)
plot(preds, studentized_residuals)
abline(h = 0, col = "red")
```

## Interpretation of residual plots for model2
-   All three plots are not showing any funnel shape, which indicates that the variance of residuals is constant across fitted values.
-   There are no points with large residuals (more than 3 standard deviations), which indicates that there are no outliers 
Compute the studentized residuals and the R ‐ student residuals for this model. What
information is conveyed by these scaled residuals?
-   Studentized residuals are the residuals divided by an estimate of their standard deviation. They are used to identify outliers in the data. 
-   R-student residuals are similar to studentized residuals but they are calculated using a different method for estimating the standard deviation. 
-   Both help in identifying outliers in the data. 
-   If there are any points with large studentized or R-student residuals (greater than 3 standard deviations), they may be considered outliers.