---
title: "Worksheet - 3"
author: "Hussain Abdul Kadhir"
date: last-modified
format:
  pdf:
    fontsize: 8pt
    fig.width: 4
    fig.height: 3
    fig.align: "center"
---
```{r}
#| include: false
knitr::opts_chunk$set(fig.width = 4, fig.height = 3)
```

Use the “cars” dataset in R. It contains speed (mph) as the independent variable and stopping distance (ft) as the dependent variable, $𝑦$: distance $𝑥$: speed Consider the two regression models:

- Model 1: $y = \beta_0 + \beta_1 x + \epsilon$

- Model 2: $y = \beta_0 + \beta_1 x + \beta_2 x^2 + \epsilon$

**Do the following:**

1.  Make a scatterplot of $𝑥$ variable (speed) versus $𝑦$ variable (distance). Are there any linear relationship between $𝑥$ and $𝑦$.

2.  Perform a log‐transformation (on $𝑥$ or $𝑦$ or both) to improve the linearity.

3.  Fit models 1 and 2 to transformed data in Part (b).

4.  Check the constant variance assumption by perfoming residual analysis.

5.  Compare Models 1 and 2 based on Adjusted $𝑅^2$. Which one is preferred?

```{r}
library(car)
data(cars)
colnames(cars)
x <- cars$speed
y <- cars$dist
# 1 scatter plot
plot(x, y)
cor(x, y)
```


## Question 1
-   Both $x$ and $y$ are showing some positive correlation, virtually linear relationship between $x$ and $y$ (correlation = 0.807)

```{r log_transform, fig.width = 5, fig.height = 4}
# 2 log transformation
x_log <- log(x)
y_log <- log(y)
plot(x_log, y_log)
cor(x_log, y_log)
plot(x, log(y))
cor(x, log(y))
plot(log(x), y)
cor(log(x), y)
```

## Question 2
-    Log transformation on $y$ and $x$ is showing the highest correlation (0.856) and better linear relationship between $x$ and $y$ compared to other transformations. So, we will use log transformation on both $x$ and $y$ for further analysis.

```{r fits, fig.width = 5, fig.height = 4}

fit1 <- lm(y ~ x)
plot(fit1$fitted.values, fit1$residuals)
abline(h = 0, col = "red")
# funnel shape observed
summary(fit1)

fit2 <- lm(log(y) ~ log(x))
plot(fit2$fitted.values, fit2$residuals)
abline(h = 0, col = "red")
# funnel shape somewhat reduced but still slightly observed
summary(fit2)

fit3 <- lm(y ~ x + I(x^2))
plot(fit3$fitted.values, fit3$residuals)
abline(h = 0, col = "red")
# funnel shape observed
summary(fit3)

fit4 <- lm(log(y) ~ x + I(log(x)^2))
plot(fit4$fitted.values, fit4$residuals)
abline(h = 0, col = "red")
# funnel shape somewhat reduced but still slightly observed
summary(fit4)
```


## Question 4
-   Summary of residual analysis for all the models:
    -   Model 1 and Model 3 are showing clear funnel shape, which indicates that the variance of residuals is not constant (heteroscedasticity).
    -   Model 2 and Model 4 are showing somewhat reduced funnel shape but still slightly observed, which indicates that the variance of residuals is somewhat improved but still not perfectly constant.


```{r summary_r2}
library(knitr)
# table of adjusted R^2 and R^2
results <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4"),
  R_squared = c(summary(fit1)$r.squared, 
                summary(fit2)$r.squared, 
                summary(fit3)$r.squared, 
                summary(fit4)$r.squared),
  Adjusted_R_squared = c(summary(fit1)$adj.r.squared, 
                          summary(fit2)$adj.r.squared, 
                          summary(fit3)$adj.r.squared, 
                          summary(fit4)$adj.r.squared)
)
kable(results)
```

## Question 5

-   summary of $R^2$and Adjusted $R^2$ for all the models:
    -   Model 2 is showing the highest $R^2$ (0.733) and Adjusted $R^2$ (0.728), which indicates that Model 2 is preferred among the four models based on these metrics.


```{r scatter_plot_with_fits, fig.width = 5, fig.height = 4}

plot(log(x), log(y), main = "Scatter plot with fitted lines", xlab = "log(speed)", ylab = "log(distance)")
abline(fit2, col = "blue", lwd = 2)

```


## Overall conclusion and interpretation

-   Based on the analysis, we can conclude that there is a positive correlation between speed and stopping distance. 
-   Log transformation on both speed and distance improved the linearity and somewhat reduced the heteroscedasticity.
-   Among the four models considered, Model 2 (log(y) ~ log(x)) is preferred based on the highest R^2 and Adjusted R^2 values.
-   This suggests that the relationship between speed and stopping distance is better captured by a log-log model compared to a simple linear or quadratic model.

